---
title: "Getting Started"
output: 
  rmarkdown::html_vignette:
    toc: true
    check_title: TRUE
vignette: >
  %\VignetteIndexEntry{Getting Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = " "
)

library(DT)
```


# What we will cover

<br>

The 6 main functions within `xportr` pacakge:

* `xportr_type()`
* `xportr_length()`
* `xportr_order()`
* `xportr_format()`
* `xportr_label()`
* `xportr_write()`


The demo will make use of a small `ADSL` dataset available with the `xportr` package and has the following features:

* 306 observations 
* 54 variables
* Data types other than character and numeric
* Missing labels on variables
* Missing label for data set
* Order of variables not following specification file
* Formats missing

<br>

```{r, eval = TRUE, message = FALSE, warning = FALSE}
library(haven)
library(dplyr)
library(labelled)
library(xportr)

data("adsl_xportr", package = "xportr")
```

<br>

```{r, echo = FALSE}
DT::datatable(adsl_xportr, options = list(
  autoWidth = FALSE, scrollX = TRUE, pageLength = 5,
  lengthMenu = c(5, 10, 15, 20)
))
```

<br>

# Preparing your Specification Files

<br>


In order to make use of the functions within `xportr` you will need to create an R data frame  that contains your specification file.  You will most likely need to do some pre-processing of your spec sheets after loading in the spec files for them to work appropriately with the `xportr` functions.  Please see our example spec sheets in `system.file(file.path("specs", "ADaM_spec.xlsx"), package = "xportr")`.

<br>

```{r}
var_spec <- readxl::read_xlsx(
  system.file(file.path("specs", "ADaM_spec.xlsx"), package = "xportr"),
  sheet = "Variables"
) %>%
  dplyr::rename(type = "Data Type") %>%
  dplyr::rename_with(tolower)
```

<br>

Below is a quick snapshot of the specification file pertaining to the `ADSL` data set, which we will make use of in the 6 `xportr` functions below.

<br>

```{r, echo = FALSE, eval = TRUE}
var_spec_view <- var_spec %>% filter(dataset == "ADSL")

DT::datatable(var_spec_view, options = list(
  autoWidth = FALSE, scrollX = TRUE, pageLength = 5,
  lengthMenu = c(5, 10, 15, 20)
))
```

<br>

# xportr_type() 

<br>

In order to be compliant with transport v5 specifications  an `xpt` file can only have two data types: character and numeric/dbl.  Currently the `ADSL` data set has chr, dbl, time, factor and date.

```{r, eval = TRUE}
look_for(adsl_xportr, details = TRUE)
```

<br>

Using `xport_type` and the supplied specification file, we can *coerce* the variables in the `ADSL` set to be either numeric or character. A message is given if variables were not coerced, this is due to the variables not being in the specification file.  

<br>

```{r, echo = TRUE}
adsl_type <- xportr_type(adsl_xportr, var_spec, domain = "ADSL", verbose = "message")
```

<br>

```{r, echo = FALSE, eval = TRUE}
look_for(adsl_type, details = TRUE)
```

# xportr_length() 

<br>

Next we can apply the lengths from a variable level specification file to the data frame.  `xportr_length` will identify variables that are missing from your specification file.  The function will also alert you to how many lengths have been applied successfully.  Before we apply the lengths lets verify that no lengths have been applied to the original dataframe.

<br>

```{r}
capture.output(str(adsl_xportr, give.head = TRUE)) %>%
  as_tibble() %>%
  head(n = 7)
```

<br>

No lengths have been applied to the variables as seen in the printout for the first 2 variables.  Let's now use `xportr_length` to apply our lengths from the specification file.

<br>

```{r}
adsl_length <- adsl_xportr %>% xportr_length(var_spec, "ADSL", "message")
```

<br>

```{r}
capture.output(str(adsl_length, give.head = TRUE)) %>%
  as_tibble() %>%
  head(n = 7)
```

<br>

Lengths have been successfully applied as viewed for the first 3 variables.

# xportr_order()

asdf

# xportr_format()

asdf

# xportr_label() 

<br>

Please observe that our `ADSL` dataset is missing many variable labels.  Sometimes these labels can be lost while using R's function.  However,  A CDISC compliant data set needs to have each variable with a variable label. 

```{r, eval = TRUE}
look_for(adsl_xportr, details = FALSE)
```

<br>

Using the `xport_label` function we can take the specifications file and label all the variables available. `xportr_label` will produce a warning message if you the variable in the data set is not in the specification file. 

<br>

```{r}
adsl_update <- adsl_xportr %>% xportr_label(var_spec, "ADSL", "message")
```

```{r}
look_for(adsl_update, details = FALSE)
```

# xportr_df_label 

<br>

An appropriate data set label must be supplied as well.  Currently, the `ADSL` data set has the label *ADSL*, but it needs the label *Subject-Level Analysis Dataset* to be compliant with most clinical data set validator application.  Here we make use of the data_spec object to supply the apropriate label for the `ADSL` data set.

#### Option1

```{r, eval = TRUE}
capture.output(str(adsl_xportr))[45]
```

```{r, eval = FALSE}
adsl_df_lbl <- adsl_xportr %>% xportr_df_label(data_spec, "ADSL")
adsl_xportr %>% xportr_varnames("message")
```

```{r, eval = FALSE}
attr(adsl_xportr, "label")
```

<br>

# xportr_varnames

<br>

This is a development feature, not currently implemented.

There are several constraints placed upon variable naming. According to modern CDISC Implementation guides and the latest [FDA conformance guide](https://www.fda.gov/media/136460/download), all ADaM variable names must be:

1. no more than 8 characters in length

1. start with a letter (not an underscore)

1. be comprised of only uppercase letters (A-Z), numerals (0-9)

1. free of non-ASCII text, symbols, or underscores. Note: Underscores are permitted only for legacy studies started on or before Dec 17th, 2016


For strings containing variable names `xportr_tidy_rename()` was designed to identify and rename variables into compliance. Thus, you can use this function to rename the spec file variables or column names on the fly. Below, we use it's default options to tidy names for submission compliance.
```{r eval=FALSE}
renamed_var_spec <- var_spec %>%
  filter(dataset == "ADSL") %>%
  mutate(tidy_variable = xportr_tidy_rename(variable)) %>%
  select(order, dataset, variable, tidy_variable, tidyselect::everything())
```

Note the above messages detail the rule(s) and variable names that were out of compliance, the number of renamed variables, and even the old and newly tidied names. The function uses a step-wise renaming algorithm to maintain the original variable names characteristics as much as possible. Below is the view of the the old variable names juxtaposed the new.

<br>

```{r, echo = FALSE, eval = FALSE}
just_new_names <-
  renamed_var_spec %>%
  filter(variable != tidy_variable)

DT::datatable(just_new_names, options = list(
  autoWidth = FALSE, scrollX = TRUE, pageLength = 5,
  lengthMenu = c(5, 10, 15, 20)
))
```

The function `xportr_varnames()` takes `xportr_tidy_rename` a step further to help users change the data.frame columns directly. It also is slightly less flexible from a customization perspective, attempting to follow the submission constraints precisely where as `xportr_tidy_rename` can be used for a wider breadth of renaming applications. Executing the code below, we create a new data.frame called `adsl_renamed`, but the variable names are already compliant. The `identical` function below shows us that nothing has changed.
```{r, eval = FALSE}
adsl_renamed <- adsl_xportr %>% xportr_varnames()
identical(adsl_xportr, adsl_renamed)
```


In the interest of showcasing `xportr_varnames` capabilities, we create a fictional data.frame `adxx` riddled with non-compliant variable names below. In fact, `adxx` only has one valid variable name: "STUDYID". Calling the function shows this data.frame's variables violate all four compliance rules and proceeds to rename all but "STUDYID".

```{r, eval = FALSE}
varnames <- c(
  "", "STUDYID", "studyid", "subject id", "1c. ENT", "1b. Eyes",
  "1d. Lungs", "1e. Heart", "year number", "1a. Skin_Desc"
)
adxx <- data.frame(matrix(0, ncol = 10, nrow = 3))
colnames(adxx) <- varnames

xportr_varnames(adxx) # default behavior
```

You may have observed in the previous output that all variables starting with a number were bundled and relocated to the END of the term. For example, 1d. Lungs" became "LUNGS1D". To maintain the numeric prefix at the start (or left-hand side), users can set a letter [A-Z] to prefix the numeric prefix. Below we set `letter_for_num_prefix` to "x" and `relo_2_end` to `FALSE`.

Second, if your organization holds an ontology of controlled terms, you can then map non-compliant variable names to any desired result using the `dict_dat` argument. It's essentially provides a "find any replace" functionality for any instance of the term(s) found in the data. In the example below, we want "subject id" to be portrayed as "SUBJID". Thus, we created a data.frame called `my_dictionary` that maps that relationship. 

```{r, eval = FALSE}
my_dictionary <- data.frame(original_varname = "subject id", dict_varname = "SUBJID")
xportr_varnames(adxx,
  relo_2_end = FALSE,
  letter_for_num_prefix = "x",
  dict_dat = my_dictionary
) # 'SUBJID' used
```

Please review the documentation using `?xportr_tidy_rename` OR `?xportr_varnames` to learn how the abbreviation algorithm works and to further customize the renaming of your variable names using a slurry of additional arguments.

<br>

# xportr_write 

<br>

Finally, we arrive at exporting the R data frame object as a xpt file with the function `xportr_write`.  The xpt file will be written directly to your current working directory.  To make it more interesting, we have put together all five function with the magrittr pipe.  A user can now coerce, apply length, variable labels, data set label and write out their final xpt file in one pipe!  Appropriate warnings and messages will be supplied to a user for any potential issues before sending off to standard clinical data set validator application or data reviewers.  

```{r, eval=FALSE}
adsl_xportr %>%
  xportr_type(var_spec, "ADSL", "message") %>%
  xportr_length(var_spec, "ADSL", "message") %>%
  xportr_label(var_spec, "ADSL", "message") %>%
  xportr_df_label(data_spec, "ADSL") %>%
  xportr_varnames("message") %>%
  xportr_write("adsl.xpt")
```

Optionally, leave out xportr_varnames and instead use `tidy_varnames = TRUE` in xportr_write() to accept the recommended defaults of `xportr_tidy_rename`.

```{r, eval=FALSE}
# No xportr_varnames()!
adsl_xportr %>%
  xportr_type(var_spec, "ADSL", "message") %>%
  xportr_length(var_spec, "ADSL", "message") %>%
  xportr_label(var_spec, "ADSL", "message") %>%
  xportr_df_label(data_spec, "ADSL") %>%
  xportr_write("adsl.xpt", tidy_varnames = TRUE)
```
