---
title: "Getting Started"
output: 
  rmarkdown::html_vignette:
    toc: true
    check_title: FALSE
vignette: >
  %\VignetteIndexEntry{xportr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = " "
)

library(DT)
```

We will demonstrate the 5 main functions within `xportr` on  an **adsl** data set that is included in this package. Other vignettes within this package make use of the CDISC Pilot Study Data.  The **adsl** data set has 24 observations and 23 variables.  

<br>

```{r, eval = TRUE, message = FALSE}
library(haven)
library(dplyr)
library(xportr)

adsl <- haven::read_sas( system.file("extdata", "adsl.sas7bdat", package="xportr"))

```

<br>

```{r, echo = FALSE}
DT::datatable(adsl, options = list(
  autoWidth = FALSE, scrollX = TRUE, pageLength = 5,
  lengthMenu = c(5, 10, 15, 20)
))
```

<br>

# Preparing your Specification Files

In order to make use of the functions within `xportr` you will need to create two R data frame objects that contain your data set specifications.  For our examples, we have referenced specifications files contained in the package.  Here we have called those two objects **var_spec** and **data_spec**.  Please note, the change of .  You will most likely need to do some pre-processing of your spec sheets after loading in the spec files.

<br>

```{r}
var_spec <- readxl::read_xlsx(
  system.file("specs", "ADaM_spec.xlsx", package="xportr"), sheet = "Variables") %>%
  dplyr::rename(type = "Data Type") %>%
  rlang::set_names(tolower)
  
data_spec <- readxl::read_xlsx(
  system.file("specs", "ADaM_spec.xlsx", package="xportr"), sheet = "Datasets") %>%
  rlang::set_names(tolower) %>%
  dplyr::rename(label = "description")

```

The spec file within ths package contains  34 of the most CDISC ADaM data sets.  Below is a quick snapshot of the specification file pertaining to the **adsl** data set, which we will make use of in the 5 `xportr` functions below.

```{r, echo = FALSE, eval = TRUE}
var_spec_view <- var_spec %>% filter(dataset == "ADSL")

DT::datatable(var_spec_view, options = list(
  autoWidth = FALSE, scrollX = TRUE, pageLength = 5,
  lengthMenu = c(5, 10, 15, 20)
))
```

# xportr_type 

```{r, eval = TRUE}
adsl %>% xportr_type(var_spec, "ADSL", "message")  
```

# xportr_length 

```{r, eval = FALSE}
adsl %>%
  xportr_type(var_spec, "ADSL", "message")  %>% 
  xportr_length(var_spec, "ADSL", "message")
```

# xportr_label 

```{r, eval = FALSE}
adsl %>% xportr_label(var_spec, "ADSL", "message") 
```

# xportr_df_label 

```{r, eval = FALSE}
adsl %>%
  xportr_type(var_spec, "ADSL", "message") %>%
  xportr_length(var_spec, "ADSL", "message")
  xportr_label(var_spec, "ADSL", "message") %>%
  xportr_df_label(data_spec, "ADSL") %>%
```
 
# xportr_write 

```{r, eval=FALSE}
adsl <- haven::read_sas("inst/extdata/adsl.sas7bdat")

var_spec <- readxl::read_xlsx("inst/specs/ADaM_spec.xlsx", sheet = "Variables") %>%
  dplyr::rename(type = "Data Type") %>%
  rlang::set_names(tolower)
  
data_spec <- readxl::read_xlsx("inst/specs/ADaM_spec.xlsx", sheet = "Datasets") %>%
  rlang::set_names(tolower) %>%
  dplyr::rename(label = "description")
  
adsl %>%
  xportr_type(var_spec, "ADSL", "message") %>%
  xportr_length(var_spec, "ADSL", "message")
  xportr_label(var_spec, "ADSL", "message") %>%
  xportr_df_label(data_spec, "ADSL") %>%
  xportr_write("adsl.xpt")
```


